Make any stupid plot you want

### 2021-02-25 draft/sections/intro.tex
halfway rework of visual section

\subsection{Visualization}
\label{sec:intro_viz}

A visualization tool produces a graphical design and an image rendered based on that design, as described by Mackinlay \cite{mackinlayAUTOMATICDESIGNGRAPHICAL1987}. He defines the graphical design as the set of encoding relations from data to visual representation\cite{mackinlayAutomatingDesignGraphical1986}, and the design rendered in an idealized abstract space is what throughout this paper we will refer to as a graphic. In addition to the graphic representations, Byrne et al. describe how visualizations have figurative representations that have meaning due to their similarity in shape to external concepts \cite{byrneAcquiredCodesMeaning2016}. Mackinlay  proposes that a visualization tool's expressiveness is a measure of how much of the structure of the data the tool encodes, while the tool's effectiveness describes how much design choices are made in deference to perceptual saliency \cite{clevelandResearchStatisticalGraphics1987,clevelandGraphicalPerceptionTheory1984,chambersGraphicalMethodsData1983a, munznerVisualizationAnalysisDesign2014}.

@@

@@

\begin{figure}[H]
\includegraphics[width=1\textwidth]{figures/intro/retinal_variables.png}
\caption{Retinal variables are a codification of how position, size, shape, color and texture are used to illustrate variations in the components of a visualization. The best to show column describes which types of information can be expressed in the corresponding visual encoding. This tabular form of Bertin's retinal variables is from Understanding Graphics \cite{malamedInformationDisplayTips2010} who reproduced it from \textit{Making Maps: A Visual Guide to Map Design for GIS} 
\cite{krygierMakingMapsVisual2005}}
\label{fig:intro_retinal_variables}
\end{figure}

The visual properties were first codified by Bertin\cite{bertinSemiologyGraphicsDiagrams2011a}, who also described the types of data that paired with the visual properties shown in figure~\ref{fig:intro_retinal_variables}. It is at this encoding level that Mackinlay formalized expressiveness as a homomorphic mapping which preserves some binary operator from one domain to another \cite{mackinlayAUTOMATICDESIGNGRAPHICAL1987}. A functional dependency framework for evaluating these equivariances was proposed by Sugibuchi et al \cite{sugibuchiFramwork2009}, and an algbraic basis for visualization design and evaluation was proposed by Kindlmann and Scheideggar\cite{kindlmann2014algebraic}. Vickers et al. propose a category theory framework\cite{vickersUnderstandingViz2013} that extends the equivariance checks to layout, but is focused strictly on the design layer like the other mathematical frameworks.

The visual encodings are composited into the point, line, and area graphical marks, as shown in figure~\ref{fig:intro_retinal_variables}. Marks can be generalized to glyphs, which are graphical objects that convey one or more attributes of the data entity mapped to it\cite{ware2019information, munznerVisualizationAnalysisDesign2014} and minimally need to be differentiable from other visual elements \cite{ziemkiewiczEmbeddingInformationVisualization2009}. As retinal variables and marks are the building blocks of most visualizations, they must be fully expressible in a framework for building visualization tools. By building visualization concepts into the core architecture, developers can incorporate assessments of the visualization,  such as as quality metrics\cite{bertiniQualityMetricsHighdimensional2011a} or invariance \cite{kindlmann2014algebraic} of visualizations with respect to graphical encoding choices.

>>>>

\subsection{Visualization}
\label{sec:intro_viz}
\begin{figure}[H]
\includegraphics[width=1\textwidth]{figures/intro/retinal_variables.png}
\caption{Retinal variables are a codification of how position, size, shape, color and texture are used to illustrate variations in the components of a visualization. The best to show column describes which types of information can be expressed in the corresponding visual encoding. This tabular form of Bertin's retinal variables is from Understanding Graphics \cite{malamedInformationDisplayTips2010} who reproduced it from \textit{Making Maps: A Visual Guide to Map Design for GIS} 
\cite{krygierMakingMapsVisual2005}}
\label{fig:intro_retinal_variables}
\end{figure}

Visual representations of data have always tried to reflect something of the underlying structure of the data\cite{friendlyBriefHistoryData2008}. The components of the visual representation were first codified by Bertin\cite{bertinSemiologyGraphicsDiagrams2011a}. As shown in figure~\ref{fig:intro_retinal_variables}, he proposes that there are classes of visual encodings such as shape, color, and texture and which type of measurement, quantitative or qualitative, will be preserved when the data is mapped to that encoding. Furthermore he proposes that the visual encodings be composited into graphical marks that match the connectivity of the data - for example discrete data is a point, 1D continuous is the line, and 2D data is the area mark. A general form of marks are glyphs, which are graphical objects that convey one or more attributes of the data entity mapped to it\cite{ware2019information, munznerVisualizationAnalysisDesign2014} and minimally need to be differentiable from other visual elements \cite{ziemkiewiczEmbeddingInformationVisualization2009}. The set of encoding relations from data to visual representation is termed the graphical design by Mackinlay \cite{aumackinlayAutomatingDesignGraphical1986} and the design rendered in an idealized abstract space is what throughout this paper we will refer to as a graphic.

A visualization tool produces a graphical design and an image rendered based on that design, as described by Mackinlay \cite{mackinlayAUTOMATICDESIGNGRAPHICAL1987}. Mackinlay  proposes that a visualization tool's expressiveness is a measure of how much of the structure of the data the tool encodes, while the tool's effectiveness describes how much design choices are made in deference to perceptual saliency \cite{clevelandResearchStatisticalGraphics1987,clevelandGraphicalPerceptionTheory1984,chambersGraphicalMethodsData1983a, munznerVisualizationAnalysisDesign2014}.

It is at this encoding level that Mackinlay formalized expressiveness as a homomorphic mapping which preserves some binary operator from one domain to another \cite{mackinlayAUTOMATICDESIGNGRAPHICAL1987}. A functional dependency framework for evaluating these equivariances was proposed by Sugibuchi et al \cite{sugibuchiFramwork2009}, and an algbraic basis for visualization design and evaluation was proposed by Kindlmann and Scheideggar\cite{kindlmann2014algebraic}. Vickers et al. propose a category theory framework\cite{vickersUnderstandingViz2013} that extends the equivariance checks to layout, but is focused strictly on the design layer like the other mathematical frameworks.

@@

In addition to the graphic representations, Byrne et al. describe how visualizations have figurative representations that have meaning due to their similarity in shape to external concepts \cite{byrneAcquiredCodesMeaning2016}.

@@

 As retinal variables and marks are the building blocks of most visualizations, they must be fully expressible in a framework for building visualization tools. By building visualization concepts into the core architecture, developers can incorporate assessments of the visualization,  such as as quality metrics\cite{bertiniQualityMetricsHighdimensional2011a} or invariance \cite{kindlmann2014algebraic} of visualizations with respect to graphical encoding choices.
